---
authors:
- admin
date: "2020-02-29T00:00:00Z"
draft: false
featured: false
image:
  caption: 'Image credit: [spacecentre](https://www.spacecentre.nz/resources/faq/solar-system/earth/img/earth-rotation.png)'
  focal_point: ""
  placement: 2
  preview_only: true
lastmod: "2019-10-01T00:00:00Z"
projects: [telematics]
subtitle: 'Rotation matrix'
summary: Introduction to the concept of rotation in 2D space.
title: 'Rotations in 2D space'
categories:
  - telematics
  - mathematics
tags:
  - Trigonometric
  - Linearalgebra
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

This article is the introduction to the basic concept in 2D space. The following are the `R` packages used in this post.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(tibble)

```

```{r message=FALSE, warning=FALSE}
library(magrittr)
library(ggplot2)
theme_set(theme_bw())
```

## Rotation of a vector in 2D plane

Let us consider a vector $\underline{a} = (3, 1)$ on the 2D plane, then we can visualize the vector as follows;

```{r echo=FALSE}
mypoint <- data.frame(x = 3, y = 1)
ggplot(data = mypoint) +
  geom_segment(aes(x = 0, y = 0, xend = x, yend = y), arrow = arrow(length = unit(0.03, "npc"))) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  ylim(0, 3) + xlim(0, 4) + coord_fixed(ratio = 1)
```

Now, if we rotate the vector $\underline{a}$ by a counterclockwise angle $30^\circ$ with respect to the origin, then the result vector $\underline{a}'$ will look like this.

```{r echo=FALSE}
rotation_mat <- function(theta, degree = FALSE){
  if (degree){
    theta <- theta * pi / 180
  }
  rmat <- matrix(c(cos(theta), -sin(theta),
                   sin(theta), cos(theta)), nrow = 2, byrow = T)
  return(rmat)
}
rmat <- rotation_mat(30, degree = TRUE)
a_prime <- as.numeric(rmat %*% as.numeric(mypoint))

ggplot(data = mypoint) +
  geom_segment(aes(x = 0, y = 0, xend = x, yend = y),
               arrow = arrow(length = unit(0.03, "npc"))) +
  geom_segment(aes(x = 0, y = 0, xend = a_prime[1], yend = a_prime[2]), 
               arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(aes(x = x, y = y, xend = a_prime[1], yend = a_prime[2]), linetype = 2, 
               arrow = arrow(length = unit(0.03, "npc"))) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  ylim(0, 3) + xlim(0, 4) + coord_fixed(ratio = 1)
```


### Rotation matrix 

In high school, we have learned this counter clockwise rotation of a vector by angle $\theta$ (radian) can be represented by the matrix $R$, called a rotation matrix, as follows:
$$
R_{\theta}=\left(\begin{array}{cc}
cos\left(\theta\right) & -sin\left(\theta\right)\\
sin\left(\theta\right) & cos\left(\theta\right)
\end{array}\right)
$$
Also, the rotation process can be written as the multiplication of the rotation vector and the given vector as follows:
$$
\underline{a}' = R_{\theta} \underline{a}
$$

Thus, in the above example, since we have a couter clockwise $30 ^\circ$ rotation, we have the following rotation matrix:
```{r}
rotation_mat <- function(theta, degree = FALSE){
  if (degree){
    theta <- theta * pi / 180
  }
  rmat <- matrix(c(cos(theta), -sin(theta),
                   sin(theta), cos(theta)), nrow = 2, byrow = T)
  return(rmat)
}
rmat <- rotation_mat(30, degree = TRUE)
rmat

```

Using this rotation matrix, we can calculate the end point of the rotated vector, $\underline{a}'$, as follows:

```{r}
as.vector(rmat %*% c(3, 1))
```

## Rotation of frame in 2D space

In the previous section, the rotated object was a vector. However, sometime we have a situation where the coordinate system is rotated rather than the vector. Let us consider the same example that we had in the previous section. We have the same vector $\underline{a}$ on the initial 2D space and we call this coordinate system as a reference frame. In this reference frame, the coordinates of the vector is $(3, 1)$, because the vector $\underline{a}$ can be represented as a linear combination of the basis vector $e_1, e_2$ of the frame as follows:
$$
\underline{a}=3e_1 + 1e_2,
$$
where $e_1$ and $e_2$ are
$$
e_1 = (1, 0)^T, \quad e_2 = (0, 1)^T
$$

### Rotating the reference frame

What if the reference frame is rotated by an clockwise angle $\theta$ to another frame? And what will be the coordinates of the vector $\underline{a}$ in the new rotated frame? The following figure shows the situation where the intial frame is rotated by a clockwise $30 ^\circ$. The solid red lines represent the x and y axis in the new frame, called $x'$ and $y'$.

```{r echo=FALSE}
rmat <- rotation_mat(-30, degree = TRUE)
p_1 <- as.numeric(rmat %*% c(1, 0))
p_2 <- as.numeric(rmat %*% c(0, 1))

rmat <- rotation_mat(-30, degree = TRUE)
p_3 <- as.numeric(rmat %*% c(0, 2.366025))
p_4 <- as.numeric(rmat %*% c(2.098076, 0))

mypoint_1 <- data.frame(x = p_1[1], y = p_1[2])
mypoint_2 <- data.frame(x = p_2[1], y = p_2[2])
ggplot(data = mypoint) +
  geom_segment(aes(x = 0, y = 0, xend = x, yend = y), arrow = arrow(length = unit(0.03, "npc"))) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_abline(slope = p_1[2] / p_1[1], intercept = 0, col = "red") +
  geom_abline(slope = p_2[2] / p_2[1], intercept = 0, col = "red") +
  annotate("text", x = 2.7, y = 3, label = "Rotated frame", col = "red") +
  annotate("text", x = 1.5, y = 3, label = "y'", col = "red") +
  annotate("text", x = 4, y = -2, label = "x'", col = "red") +
  annotate("text", x = 3.2, y = 1.2, label = "(2.098076, \n                2.366025)", col = "red") +
  annotate("text", x = 0.7, y = 3, label = "Initial frame", col = "black") +
  geom_segment(aes(x = 3, y = 1, xend = p_3[1], yend = p_3[2]), linetype = "dashed", col = "red") +
  geom_segment(aes(x = 3, y = 1, xend = p_4[1], yend = p_4[2]), linetype = "dashed", col = "red") +
  ylim(-2, 3) + xlim(0, 4) + coord_fixed(ratio = 1)
```
In this new frame, the coordinates of the vector $\underline{a}$ will be $(2.098076, 2.366025)$, which is the same coordinates we had in the previous vector rotation case. Thus, rotating the initial frame by a clockwise angle $\theta$ has the same effect as the counter clockwise rotation of the vector by an angle $\theta$ when we consider the measurements of the vector. 

Let us define some notations to describe the situation more precisely. Let us denote $I$ and $B$ by the initial frame and the rotated frame (usually called body frame) respectively. Then, the rotation matrix $R^{B|I}_{\theta}$ can be defined as a rotation matrix, whose angle $\theta$ represents the angle between the bodyframe and the initial frame where the initial frame is the reference for measuring angle $\theta$. Note that the angle $\theta$ is measured under the right-hand rule. In our example, a counter clockwise corresponds to the positive rotation while the clockwise corresponds to the negative rotation. Then, we can set up the relationship between the measurement of a given vector $\underline{a}$ in the initial frame and the body frame as follows:
$$
\underline{a}_{B} = R^{B|I}_{\theta} \underline{a}_{I}
$$
where $p_I$ and $p_B$ represent the measurement of the vector $\underline{a}$ in the initial frame and the body frame



### Reference

[1] Luke Tierney, STAT:7400 Computer Intensive Statistics Note:
http://homepage.divms.uiowa.edu/~luke/classes/STAT7400/notes.pdf
  
